### 2013-01-15 paper_sean/conclusion.tex
reverted back to means/std

With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 54.3\% in simulation and 5.1\% in reality over the previous best QuadraTot gait by Glette et al. These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for a legged robot. The results further reaffirm the benefits of using simulators when solving real-world challenges with evolutionary algorithms. Our results additionally confirm---with a different robot and simulator---previous work that has shown that HyperNEAT is an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}. That HyperNEAT outperformed the encoding hand-designed by Glette et al. shows that HyperNEAT can outperform even evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial for a problem. The results thus demonstrate that automatically discovering regularities can be a superior approach to specifying them, even on problems that are relatively well-understood.

>>>>

With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 52.1\% in simulation and 5.1\% in reality over the previous best QuadraTot gait by Glette et al. These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for a legged robot. The results further reaffirm the benefits of using simulators when solving real-world challenges with evolutionary algorithms. Our results additionally confirm---with a different robot and simulator---previous work that has shown that HyperNEAT is an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{yos:clune,clune2011performance,clune2009evolving}. That HyperNEAT outperformed the encoding hand-designed by Glette et al. shows that HyperNEAT can outperform even evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial for a problem. The results thus demonstrate that automatically discovering regularities can be a superior approach to specifying them, even on problems that are relatively well-understood.

### 2013-01-15 paper_sean/introduction.tex
reverted back to means/std

Legged robots can operate in a much wider range of environments than their wheeled counterparts. However, designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a robot is created or modified~\cite{hornby2005autonomous}. 
Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, and the result is often a better gait than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}. While it has been shown that gaits perform better if they are \emph{regular}---i.e., that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}---experimenters usually have to explicitly decide and specify these regularities~\cite{valsalam:mii,tellez,beer,raibert}. 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.

Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits that outperform gaits evolved with direct encodings~\cite{clune2009evolving,clune2011performance}.
However, that work only verified these claims in simulation. 
Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform~(\figref{quadratotWhiteBooties}). They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.\footnote{Videos available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}
A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in simulation with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}. 
The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware, resulting in 333 times more evaluations per run (60000 vs. 180)~\cite{yos:clune,glette}.

The work with the simulator~\cite{glette} evolved gaits with a simple encoding manually constrained to produce specific regularities. The success of that work raises the question of whether the performance gains were due to the added benefit of a simulator or the use of a simple, hand-designed encoding. We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al. if combined with a simulator. 
Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al. and then transferring those gaits to the QuadraTot robot.  Our experiments confirmed the hypothesis: with the simulator, HyperNEAT evolved the highest-performing gaits observed to date for the QuadraTot platform.

>>>>

Legged robots can operate in a much wider range of environments than their wheeled counterparts. However, designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a robot is created or modified~\cite{hornby2005autonomous}. 
Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, and the result is often a better gait than those created by human engineers~\cite{hornby2005autonomous,valsalam:mii,kohl:stone,hornby2003generative,yos:clune}. While it has been shown that gaits perform better if they are \emph{regular}---i.e., that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}---experimenters usually have to explicitly decide and specify these regularities~\cite{valsalam:mii,tellez,beer,raibert}. 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.

Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits that outperform gaits evolved with direct encodings~\cite{clune2011performance,clune2009evolving}.
However, that work only verified these claims in simulation. 
Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform~(\figref{quadratotWhiteBooties}). They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.\footnote{Videos available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}
A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in simulation with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}. 
The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware, resulting in 333 times more evaluations per run (60000 vs. 180)~\cite{yos:clune,glette}.

The work with the simulator~\cite{glette} evolved gaits with a simple encoding manually constrained to produce specific regularities. The success of that work raises the question of whether the performance gains were due to the added benefit of a simulator or the use of a simple, hand-designed encoding. We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{yos:clune,clune2011performance}, would outperform the simpler encoding from Glette et al. if combined with a simulator. 
Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al. and then transferring those gaits to the QuadraTot robot.  Our experiments confirmed the hypothesis: with the simulator, HyperNEAT evolved the highest-performing gaits observed to date for the QuadraTot platform.

### 2013-01-15 paper_sean/results.tex
reverted back to means/std

%We
%compare performance in simulation and hardware
%to available data from previous studies.
%\subsubsection{Simulation:}

The simulator enabled HyperNEAT to evolve
fast, organic gaits. In simulation, HyperNEAT gaits were significantly faster than those from Glette et al. (\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).  
Specifically, HyperNEAT gaits were 61.1\% faster in simulation (26.9
cm/s, 95\% CIs 23.8 cm/s and 26.750 cm/s versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).
Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.

\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.00}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
{
\textbf{Top: }
HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator. Plotted are medians over 20 runs in simulation (solid lines) $\pm$ CIs (dashed lines). HyperNEAT gaits are 61.1\% faster in
simulation and 5.1\% faster in reality~(\tabref{results}).
\textbf{Left: }
Servo positions over time (for nine servos) for a representative simulated HyperNEAT gait. HyperNEAT produced smooth and symmetrical gaits that contained complex regularities. 
\textbf{Right: }Mean gait frequency averaged over 20 runs. 
Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learned to produce gaits with frequencies low enough to avoid this penalty.
}

On the physical robot, HyperNEAT gaits from this study outperformed gaits from all previous QuadraTot studies~(\tabref{results})~\cite{yos:clune,haocheng,glette}, including those of Glette et al. 
 However, comparing performance in hardware between studies performed in
different laboratories is difficult, not only because reality is
inherently noisy, but because even two copies of the same robot are
not identical and may produce different speeds for the same input
gait. Gaits for this study and two previous studies~\cite{yos:clune,haocheng} were evaluated on a copy of the QuadraTot robot in the Cornell Creative Machines Lab (CCML), while the gaits in Glette et al. were evaluated on a different copy of the same robot in the Robotics and
Intelligent Systems (ROBIN) lab at the University of Oslo. The two robots were evaluated on two different surfaces, and had different material enveloping their feet to increase friction (compare \figref{quadratotWhiteBooties}-Left to Figure 2 of Glette et al).

@@

\begin{table}
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline
                                         &              & Simulated  & Real Vel. & Real Vel.  \\
                                         & Evaluations  & Velocity &    (CCML)     & (ROBIN) \\
\hline
Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 5.8 & --\\
\hline
HyperNEAT in hardware \cite{yos:clune}                 & 180         & --         &   9.7  & --   \\
\hline
RL PoWER Spline \cite{haocheng}                         & 300         & --         &   11.1 & --\\
\hline
GA + simulator \cite{glette}             & 60000       & *16.7       &   13.8   & \textbf{17.8}  \\
\hline
HyperNEAT + simulator [this paper]                     & 40000       & \textbf{26.9}       &   \textbf{14.5} & --\\
\hline
\end{tabular}
\vspace{.35cm}
\caption{Velocities of evolved gaits in simulation and on two different copies of the QuadraTot robot. Subject to availability, data are reported from the experiments for this paper and three previous studies. Reported are the total number of evaluations per run, the median of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences). *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs. Velocities are in cm/s, and bold indicates the best performance. }  
\tablabel{results}
\end{center}
\end{table}

@@

The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population size (200 vs. 9) and the number of generations (200 vs. 20), leading
to a total difference of 40000 vs 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
potential cause of improved performance is that with a simulator
HyperNEAT had much less noise in the evaluation process and could thus find coordinated, regular, gaits, which perform better. On the physical robot, the noise in the evaluation was
substantial, preventing effective learning~\cite{yos:clune}. 
However, we can likely rule out the noise hypothesis because we performed 20 
runs in simulation with the same evaluation samples as 
in Yosinski et al.~\cite{yos:clune} and saw that the results were not 
significantly different ($p < 0.855$, mean fitness 7.9 cm/s), making the likely explanation 
the number of evaluations performed.
The encouragement of low-frequency gaits in this study
also may have aided performance, especially since in simulation the gaits were
high-frequency in a few early generations before rapidly settling to a
 range below the penalized threshold~(\figref{threeplots}-Right).

While this study was able to produce the fastest QuadraTot gait to
date, most of the gaits in simulation did not transfer well to
reality. Many gaits that were fast in simulation performed
poorly on the real robot, largely due to servos that were too weak and
shut down, or because of differences between simulation and
reality. Repeated attempts to minimize these problems were
unsuccessful.  A major, unavoidable cause of the servo shutdowns was
the design of the robot: the force required to lift the robot in many
poses was too much for the servos. In future studies we will use a
robot that has more mechanical advantage and requires less torque from
each servo, such as the Aracna platform~\cite{lohmann2012aracna},
which was designed in response to the challenges experienced with the
QuadraTot platform. That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models. In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.

>>>>

%We%compare performance in simulation and hardware
%to available data from previous studies.
%\subsubsection{Simulation:}

The simulator enabled HyperNEAT to evolve
fast, organic gaits. In simulation, HyperNEAT gaits were significantly faster than those from Glette et al. (\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).  
Specifically, HyperNEAT gaits were 52.1\% faster in simulation (25.4
cm/s, $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results} -- qualitative conclusions remain the same even if the analysis is conducted using medians, but to make it easier for comparisons with earlier works~\cite{glette}, we use means $\pm$ SD).
Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.

%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
\figOneOnTwo[h!]{threeplots}{avg_fit_vs_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.00}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
{
\textbf{Top: }
%HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator. Plotted are medians over 20 runs in simulation (solid lines) $\pm$ CIs (dashed lines). HyperNEAT gaits are 52.1\% faster in
HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator. Plotted are means over 20 runs in simulation (solid lines) $\pm$ SD (dashed lines). HyperNEAT gaits are 52.1\% faster in
simulation and 5.1\% faster in reality~(\tabref{results}).
\textbf{Left: }
Servo positions over time (for nine servos) for a representative simulated HyperNEAT gait. HyperNEAT produced smooth and symmetrical gaits that contained complex regularities. 
\textbf{Right: }Mean gait frequency averaged over 20 runs. 
Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learned to produce gaits with frequencies low enough to avoid this penalty.
}

On the physical robot, HyperNEAT gaits from this study outperformed gaits from all previous QuadraTot studies~(\tabref{results})~\cite{yos:clune,glette,haocheng}, including those of Glette et al. 
 However, comparing performance in hardware between studies performed in
different laboratories is difficult, not only because reality is
inherently noisy, but because even two copies of the same robot are
not identical and may produce different speeds for the same input
gait. Gaits for this study and two previous studies~\cite{yos:clune,haocheng} were evaluated on a copy of the QuadraTot robot in the Cornell Creative Machines Lab (CCML), while the gaits in Glette et al. were evaluated on a different copy of the same robot in the Robotics and
Intelligent Systems (ROBIN) lab at the University of Oslo. The two robots were evaluated on two different surfaces, and had different material enveloping their feet to increase friction (compare \figref{quadratotWhiteBooties}-Left to Figure 2 of Glette et al).

@@

\begin{table}
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline
                                         &              & Simulated  & Real Vel. & Real Vel.  \\
                                         & Evaluations  & Velocity &    (CCML)     & (ROBIN) \\
\hline
Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 5.8 & --\\
\hline
HyperNEAT in hardware \cite{yos:clune}                 & 180         & --         &   9.7  & --   \\
\hline
RL PoWER Spline \cite{haocheng}                         & 300         & --         &   11.1 & --\\
\hline
GA + simulator \cite{glette}             & 60000       & *16.7       &   13.8   & \textbf{17.8}  \\
\hline
HyperNEAT + simulator [this paper]                     & 40000       & \textbf{**25.4}       &   \textbf{14.5} & --\\
\hline
\end{tabular}
\vspace{.35cm}
\caption{Velocities of evolved gaits in simulation and on two different copies of the QuadraTot robot. Subject to availability, data are reported from the experiments for this paper and three previous studies. Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences). *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs. Velocities are in cm/s, and bold indicates the best performance. **Median fitnesses of this paper: 26.9 cm/s with 95\% confidence intervals 23.8 cm/s and 26.75 cm/s.}  
\tablabel{results}
\end{center}
\end{table}

@@

The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population size (200 vs. 9) and the number of generations (200 vs. 20), leading
to a total difference of 40000 vs 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
potential cause of improved performance is that with a simulator
HyperNEAT had much less noise in the evaluation process and could thus find coordinated, regular, gaits, which perform better. On the physical robot, the noise in the evaluation was
substantial, preventing effective learning~\cite{yos:clune}. 
We performed 20 runs in simulation with the same evaluation samples as 
in Yosinski et al.~\cite{yos:clune} and saw that simulated gaits performed
slightly better than the hardwared learned gaits ($p < 0.1571$, mean fitness 7.9 cm/s).
This supports the noise hypothesis, but suggests that the number of evaluations performed is a more likely explanation.
The encouragement of low-frequency gaits in this study
also may have aided performance, especially since in simulation the gaits were
high-frequency in a few early generations before rapidly settling to a
 range below the penalized threshold~(\figref{threeplots}-Right).

While this study was able to produce the fastest QuadraTot gait to
date, most of the gaits in simulation did not transfer well to
reality. Many gaits that were fast in simulation performed
poorly on the real robot, largely due to servos that were too weak and
shut down, or because of differences between simulation and
reality. Repeated attempts to minimize these problems were
unsuccessful. 
%A major, unavoidable cause of the servo shutdowns was
%the design of the robot: the force required to lift the robot in many
%poses was too much for the servos. 
In future studies we will use a
robot that has more mechanical advantage and requires less torque from
each servo, such as the Aracna platform~\cite{lohmann2012aracna}.
%which was designed in response to the challenges experienced with the
%QuadraTot platform. 
That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models. In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.
